---
title: "data.table project"
author: "Eugenia Olari"
date: "21.03.2026"
format: html
---

## 1. Load data adn convert to data.table

```{r}
## load data
dta <- readRDS(file = "data/prec_data.rds")

## load package
library(data.table)

## convert to data.table
dta_t <- as.data.table(x = dta)
print(head(dta_t))
```



## 2. Basic inspection

```{r}
## check classes
class(x = dta)
class(x = dta_t)
```

```{r}
## show first rows
head(x = dta_t)
```

```{r}
## object sizes
object.size(x = dta)
object.size(x = dta_t)
```

```{r}
## column names
names(x = dta_t)
```

```{r}
## dimensions - number of rows and columns
nrow(x = dta_t)
ncol(x = dta_t)
```

```{r}
## column types
str(object = dta_t)
```

```{r}
## check for missing datetimes
sum(is.na(x = dta_t$DT))
```

```{r}
## check QUALITY values
table(dta_t$QUALITY, useNA = "always")
```

```{r}
## check FLAG values
table(dta_t$FLAG, useNA = "always")
```

Based on the inspection above:

- **Identifier variables**: `STATION`, `DT` (datetime), `ELEMENT` — these uniquely identify
each measurement. `STATION` has 431 levels, `ELEMENT` has only one level (`SRA1H`).

- **Categorical variables**: `STATION` and `ELEMENT` are factors. `FLAG` is character,
`QUALITY` is numeric.

- **Variables needing cleaning**: `X` is entirely `NA` and will be dropped. `FLAG` has
136,458 empty strings and 56,382 `"Z"` codes worth checking. `QUALITY` has 128,245 rows
with code `3` and 37,002 with code `4`, indicating lower quality measurements.




## 3. Filtering and summaries

```{r}
## filter VALUE > 0
nz <- dta_t[VALUE > 0, .(STATION, DT, VALUE, ELEMENT, FLAG, QUALITY)]
```

```{r}
## summary by station
nz[, .(mean = mean(x = VALUE),
       sd   = sd(x = VALUE),
       iqr  = IQR(x = VALUE),
       min  = min(x = VALUE),
       max  = max(x = VALUE)),
   by = STATION][order(-mean)]
```

```{r}
## summary by station and year of DT
nz[, .(mean = mean(x = VALUE),
       sd   = sd(x = VALUE),
       iqr  = IQR(x = VALUE),
       min  = min(x = VALUE),
       max  = max(x = VALUE)),
   by = .(STATION, year(x = DT))][order(-mean)] # do the grouped summary
```

The station-level summary covers 431 stations with non-zero precipitation records. 
The highest mean is at station `0-20000-0-11787` (1.04 mm/hr), the lowest at `0-20000-0-11464` (0.56 mm/hr). 
The maximum single-hour value of **64.0 mm** at station `0-203-0-11721` is a clear outlier.

When grouped by year, station `0-20000-0-11787` in **2020** has the highest mean (1.33mm/hr). 
Station `0-203-0-11721` in 2024 has the largest standard deviation (3.50), driven by the 64.0 mm extreme value.




## 4. Duplicate analysis

```{r}
## count exact duplicate rows
sum(duplicated(x = dta_t))
```
The dataset contains **2846 exact duplicate rows**.

```{r}
## show examples of exact duplicates
dta_t[duplicated(x = dta_t) | duplicated(x = dta_t, fromLast = TRUE)][1:10]
```
All duplicates share the timestamp `2018-03-25 01:00:00` — a daylight saving time clock change date, where 01:00 occurs twice. 
Since both rows in each pair have identical values, removing one copy with `unique()` is safe.

```{r}
## define logical duplicate: same station, dt, and element
## count logical duplicates using data.table
dt_dup <- dta_t[, .(n = .N),
                by = .(STATION, DT, ELEMENT)]
```

```{r}
## how many key combinations appear more than once
dt_dup[n > 1, .N]
```

```{r}
## show examples
dt_dup[n > 1][order(-n)][1:10]
```

There are **2922 logical duplicates** — `STATION`, `DT`, `ELEMENT` combinations appearing
more than once. All have `n = 2` and all share the same DST timestamp.



## 5. Cleaning using :=

```{r}
## drop stray X column
dta_t[, X := NULL]
```
The X column contained only NA values across all 25.7 million rows and carries no information


```{r}
## drop rows where DT is NA
dta_t <- dta_t[!is.na(x = DT)]
sum(is.na(x = dta_t$DT))
```
No `NA` values exist in `DT` 

```{r}
## flag suspicious values: negative precipitation is physically impossible
dta_t[, suspicious := VALUE < 0]
dta_t[suspicious == TRUE, .N]
```
No negative `VALUE` entries found 

```{r}
## flag low quality rows
dta_t[, low_quality := QUALITY %in% c(3, 4)]
dta_t[low_quality == TRUE, .N]
```
**165247 rows** have `QUALITY` code `3` or `4` and are flagged as `low_quality`. 
These rows are kept but we will kepp them in mind for cautious use.

```{r}
## remove exact duplicates
dta_t <- unique(x = dta_t)
nrow(x = dta_t)
```
After deduplication the dataset contains **25725335 rows**, down from 25728181 - a reduction of 2846 duplicate rows.


## 6. Renaming and column work

```{r}
## rename DT to DATETIME for clarity
setnames(x = dta_t,
         old = c("DT"),
         new = c("DATETIME"),
         skip_absent = TRUE)

## create year column — useful for temporal grouping
dta_t[, yr := year(x = DATETIME)]

## create row id within each station
dta_t[, row_id := 1:.N,
      by = STATION]

## remove row_id helper column after inspection
dta_t[, row_id := NULL]
names(x = dta_t)
```

`DT` is renamed to `DATETIME` for clarity using `setnames()`. 
A `yr` column is added for fast year-level grouping. 
A `row_id` helper column is created within each station and immediately removed, it was used only to verify row ordering.



## 7.Station summary

```{r}
## station summary
st_sum <- dta_t[, .(
  n_rows = .N,
  mean_val = mean(x = VALUE, na.rm = TRUE),
  sd_val = sd(x = VALUE, na.rm = TRUE),
  min_val = min(x = VALUE, na.rm = TRUE),
  max_val = max(x = VALUE, na.rm = TRUE),
  n_nonzero = sum(VALUE > 0, na.rm = TRUE),
  n_suspicious = sum(suspicious == TRUE, na.rm = TRUE),
  n_flagged = sum(!is.na(x = FLAG) & FLAG != "", na.rm = TRUE),
  n_low_qual = sum(low_quality == TRUE, na.rm = TRUE)
), by = STATION]
```

```{r}
## Q1: most problematic (most flagged + low quality)
st_sum[order(-n_flagged)][1:5]

## Q2: highest non-zero precipitation
st_sum[order(-mean_val)][1:5]

## Q3: most records
st_sum[order(-n_rows)][1:5]

## Q4: most suspicious rows
st_sum[order(-n_suspicious)][1:5]

```