---
title: "data.table project"
author: "Eugenia Olari"
date: "21.03.2026"
format: html
---

## 1. Load data adn convert to data.table

```{r}
## load data
dta <- readRDS(file = "data/prec_data.rds")

## load package
library(data.table)

## convert to data.table
dta_t <- as.data.table(x = dta)
print(head(dta_t))
```



## 2. Basic inspection

```{r}
## check classes
class(x = dta)
class(x = dta_t)
```

```{r}
## show first rows
head(x = dta_t)
```

```{r}
## object sizes
object.size(x = dta)
object.size(x = dta_t)
```

```{r}
## column names
names(x = dta_t)
```

```{r}
## dimensions - number of rows and columns
nrow(x = dta_t)
ncol(x = dta_t)
```

```{r}
## column types
str(object = dta_t)
```

```{r}
## check for missing datetimes
sum(is.na(x = dta_t$DT))
```

```{r}
## check QUALITY values
table(dta_t$QUALITY, useNA = "always")
```

```{r}
## check FLAG values
table(dta_t$FLAG, useNA = "always")
```

Based on the inspection above:

- **Identifier variables**: `STATION`, `DT` (datetime), `ELEMENT` — these uniquely identify a measurement. `STATION` is a factor with 434 levels, `ELEMENT` has only one level (`SRA1H`), meaning the entire dataset contains a single measurement type.
- **Categorical variables**: `STATION` and `ELEMENT` are factors. `FLAG` is character and `QUALITY` is numeric — both carry quality-related metadata.
- **Variables needing cleaning**: `VALUE` (dominated by valid zeros representing no rainfall, but may contain negatives or outliers), `FLAG` (mostly `NA` meaning unflagged — but 47,732 rows have an empty string `""` and 15,172 have code `"Z"`, both worth investigating), `QUALITY` (numeric — `0` likely means good quality for the majority of rows, but 34,638 rows have code `3` and 18,128 have code `4`, suggesting degraded or suspect measurements), `DT` (contains `NA` values — rows with missing timestamps cannot be placed on the time axis and must be removed), and `X` (entirely `NA` — a stray column likely from a CSV export artifact, to be dropped during cleaning).





## 3. Filtering and summaries

```{r}
## filter VALUE > 0
nz <- dta_t[VALUE > 0, .(STATION, DT, VALUE, ELEMENT, FLAG, QUALITY)]
```

```{r}
## summary by station
nz[, .(mean = mean(x = VALUE),
       sd   = sd(x = VALUE),
       iqr  = IQR(x = VALUE),
       min  = min(x = VALUE),
       max  = max(x = VALUE)),
   by = STATION][order(-mean)]
```

```{r}
## summary by station and year of DT
nz[, .(mean = mean(x = VALUE),
       sd   = sd(x = VALUE),
       iqr  = IQR(x = VALUE),
       min  = min(x = VALUE),
       max  = max(x = VALUE)),
   by = .(STATION, year(x = DT))][order(-mean)] # do the grouped summary
```

After filtering to non-zero values, 1063904 rows remain across 433 stations. The station-level summary shows notable variation in mean hourly precipitation. 

The **highest mean** is at station 0-203-0-11640 (1.15 mm/hr), followed by 0-20000-0-11787 (1.10) and 0-203-0-11785 (1.07). 

At the **lower end**, stations 0-203-0-11735 (0.42) and 0-203-0-11301156001 (0.39) show much lower means and small standard deviations. 

The **maximum single-hour** value of 64.0 mm at station 0-203-0-11721 stands out as a likely outlier — it is more than 20 standard deviations above that station's mean of 1.07.

When grouped by station and year, the **highest means are concentrated** in 2024. 

Station 0-203-0-11640 leads again with a mean of 1.26 mm/hr, while 0-203-0-11721 shows the largest standard deviation (3.50) in any station-year combination — driven by that same extreme 64.0 mm value. 

This reinforces 2024 as a year worth extra scrutiny.




## 4. Duplicate analysis

```{r}
## count exact duplicate rows
sum(duplicated(x = dta_t))
```
The dataset contains 417 exact duplicate rows — rows that are identical across all 7 columns.

```{r}
## show examples of exact duplicates
dta_t[duplicated(x = dta_t) | duplicated(x = dta_t, fromLast = TRUE)][1:10]
```
Inspecting the duplicated rows shows us that all of them have DT = NA — the timestamp is missing entirely. These are not cases of the same hour being recorded twice, but rather rows where the datetime failed to parse at the very beginning of the rds file creation, I am not sure.

```{r}
## define logical duplicate: same station, dt, and element
## count logical duplicates using data.table
dt_dup <- dta_t[, .(n = .N),
                by = .(STATION, DT, ELEMENT)]
```

```{r}
## how many key combinations appear more than once
dt_dup[n > 1, .N]
```
There are 430 logical duplicates — unique combinations of STATION, DT, and ELEMENT that appear more than once. This is slightly higher than the exact duplicate count because some NA-timestamp rows differ in other columns while still sharing the same missing key.

```{r}
## show examples
dt_dup[n > 1][order(-n)][1:10]
```

All logical duplicates have n = 2 — no combination appears more than twice. Every case involves DT = NA, confirming that the real root cause is a datetime parsing failure when the rds file was created, or even earlier. These rows will be removed in the cleaning step by dropping all rows where DT is NA.



## 5. Cleaning

```{r}
## drop stray X column
dta_t[, X := NULL]
```
The X column contained only NA values across all 7.5 million rows and carries no information


```{r}
## drop rows where DT is NA
dta_t <- dta_t[!is.na(x = DT)]
sum(is.na(x = dta_t$DT))
```
All 861 rows with missing timestamps have been removed. No NA values remain in DT

```{r}
## flag suspicious values: negative precipitation is physically impossible
dta_t[, suspicious := VALUE < 0]
dta_t[suspicious == TRUE, .N]
```
No negative VALUE entries exist in the dataset — precipitation readings are all non-negative, which is physically consistent. 

```{r}
## flag low quality rows
dta_t[, low_quality := QUALITY %in% c(3, 4)]
dta_t[low_quality == TRUE, .N]
```
These number of rows with low quality codes (3 or 4) is 52764. These rows are flagged but not removed, as they may still contain usable information depending on the analysis context.

```{r}
## remove exact duplicates
dta_t <- unique(x = dta_t)
nrow(x = dta_t)
```
After removing exact duplicates, the dataset now contains 7553236 rows — a reduction of 417 rows from the original 7500000. The logical duplicates with NA timestamps have been removed in the previous step when we dropped rows with NA in DT.


## 6. Renaming and column work